A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios

### 2015-11-09 oneFileTON.tex
Finished

\begin{itemize}
	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
	\item The impact of imperfect clocks and channel errors on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.
	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
	\item Introduce the Schedule Reset mechanism for reducing the schedule length in case of users withdrawing from the contention (non-saturated traffic), or when its value is increased due to channel errors.
	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
	\item First implementation of CSMA/ECA with Hysteresis in real hardware and the experimental results in a crowded WLAN testbed.
\end{itemize}

@@

@@

Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like LTE~\cite{canoLTEcoexistence}. Applying the same concept to CSMA networks by modifying DCF's random backoff proceedure provides similar benefits~\cite{HE}. The following are MAC protocols for WLANs, decentralised and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques. A survey of collisio-free MAC protocols for WLANs is presented in~\cite{L_MAC}. In this paper we only overview those that are similar to CSMA/ECA.

\subsection{Zero Collision MAC}

@@

ZC-MAC is able to outperform CSMA/CA under different scenarios. Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $N>M$. Furthermore, if $M$ is overestimated ($M\gg N$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance. Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends. This is not considered in CSMA/CA and constitutes an obstacle towards standardisation.

@@

\begin{figure*}[tb]
\centering
  \includegraphics[width=0.8\linewidth]{figures/basicECA.eps}
  \caption{An example of the temporal evolution of CSMA/ECA in saturation}
  \label{fig:BECA-example}
\end{figure*}

@@

Collision slots being orders of magnitude larger than empty slots degrade the network performance. When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's. Figure~\ref{fig:BECA}, shows the average aggregated throughput of CSMA/ECA, CSMA/CA and two of the collision-free protocolos presented in Section~\ref{relatedWork}, namely L-ZC and L-MAC~\cite{L_MAC}.

@@

Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders is greater than $B_{\text{d}}+1$ ($N=8$ in the case of the figure). Beyond this point, the network will have a mixed behaviour relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff. As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's. This effect is also observed in L-ZC and L-MAC. It happens becasue the virtual cicle used by these protocolos ($M$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in lower than the number of contenders, $N$.

@@

	Channel errors, hidden-nodes, or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations's backoff stages to the maximum value very quickly (contributing to the delay increase). This issue is leveraged with the Schedule Reset mechanim, introduced in Section~\ref{schedReset}.

@@

	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
	Fair Share is an A-MPDU aggregation mechanism~\cite{A-MPDU} that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness. However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads. Furthermore, the level of aggregation provided by Fair Share depends only on the buffer occupancy.

@@

	\begin{equation}
			S =
				\begin{cases}\label{eq:bound}
					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{\text{d}} \\
					Ns_{k}(l_{k}), & \text{otherwise}
				\end{cases}
	\end{equation}

@@

	\subsection{Hidden terminals}\label{hidden}
	One key characteristic of IEEE 802.11 devices is that their carrier sense range is at least two times greater than their data range~\cite{tuningCarrierSense}. In this situation, it could be expected that the impact of hidden nodes is low, as a given transmission can be only interfered by other transmissions from very distant nodes with energy levels not higher than the noise floor, which can be easily ignored. However, in specific deployments, where obstacles play also an important role on the propagation effects, hidden nodes may appear, causing asynchronous packet collisions~\cite{throughputUnderHiddenTerm}.

	In the case of CSMA/ECA$_{\text{Hys+FS}}$ WLANs, the same negative effects as in IEEE 802.11 WLANs are expected. In addition, collisions with hidden terminals will cause nodes to leave from any collision-free schedule, as channel errors do, thus continuously disrupting any attempt of collision-free operation.

@@

	Figure~\ref{fig:scheduleReset1} shows the slots between the transmissions of a contender with $B_{\text{d}}=31$. Starting from the left, the current $B_{\text{d}}=0$ means that the first slot will be filled with the contender's own transmission. Each following slot containing either a transmission or a collision is identified with the number one, while empty slots are marked with a zero. Notice that the highlighted empty slots appear every eight slots, suggesting that a schedule reduction from $B_{\text{d}}=31$ to $B_{\text{d}}^{*}=7$ is possible; where $B_{\text{d}}^{*}$ is the new deterministic backoff assigned by Schedule Reset\footnote{$B_{\text{d}}=2^{0}CW_{\min}/2-1=7$ (see Table~\ref{tab:mac-params}). So the change is made to the backoff stage, $k$. In this case from $2$, to $k=0$. This means that any new schedule must also be a power of two minus one.}. The schedule change is performed after the contender's next successful transmission.

	%, coupled with an increase by one in the node's \emph{stickiness}~\cite{L_MAC2}. 
	%\footnote{Trailing waiting periods, like the SIFS, are composed of empty slots. These allow nodes to decrement their backoff counters even after a collision or successful slots. Without loss of generality, the stations in our examples and simulations decrement their respective backoff counters by one after these kind of slots (see example in Figures~\ref{fig:ECA} and~\ref{fig:ecaQoS}).}
		\begin{figure}[tb]
		\centering
			\includegraphics[width=\linewidth]{figures/tonFigs/scheduleReset.pdf}
			\caption{Example of the Schedule Reset mechanism. The lower row shows values of $t$ for a bitmap $b[t]$ of size $B_{\text{d}}=31$}
			\label{fig:scheduleReset1}
		\end{figure}

			\begin{algorithm}[tb]
		\If{sxTx == 1}{
			Initialising\;
			\If{B == 0}{
				$b[B_{\text{d}}] = \{0\}$\;
				$t = 0$\;
			}
		}
		\If{(sxTx $>$ 0) \&\& (sxTx $\leq~\gamma$)}{
			Filling the bitmap\;
			$b[t] \|\sigma_{i}(t)$\label{bitwiseOR}\;
			$t++$\;
			\If{$t == B_{\text{d}} -1$}{
				$t=0$\;
			}
		}
		\If{sxTx == $\gamma$}{\label{evaluations}
			Analising\;
			sxTx = $0$\;
			\For{($j=0;~j<k;~j++$)}{\label{sets}
				$y=ceil(2^{j}CW_{\min}/2)$\;
				isItPossible $= 0$\;
				\For{($m=y;~m<B_{\text{d}};~m+=y$)}{
					isItPossible$~+=b[m]$\;\label{filling}
				}
				\If{isItPossible == $0$}{\label{testFilling}
					change = $1$\;
					newStage = $m$\;
					{\bf break};
				}
			}
		}
		\If{change == $1$}{
			Making the change\;
			$k$ = newStage\;
			change = $0$\;\label{endAssignment}
		}
		\vspace{0.2cm}
		\caption{Schedule Reset Mechanism for CSMA/ECA$_{\text{Hys+FS}}$. Every consecutive successful transmission increases the variable SxTx by one, while a collision resets it to zero. The algorithm is called a \emph{reset} when all smaller deterministic backoff are tested. On the other hand, when $j$ in line~\ref{sets} is initialised to $j=k-1$, it is called a schedule \emph{halving}}\label{alg:schedRest}
	\end{algorithm}

@@

	\subsection{Channel errors}\label{channelErrorsDef}
	Channel errors are modelled by assigning a probability of a packet being corrupted by the channel, $p_e>0$. That is, in a single packet transmission there is probability $p_e$ that the transmission will not be acknowledged. If the transmission is an A-MPDU (like in the case of CSMA/ECA$_{\text{Hys+FS}}$), $p_e$ will affect each MPDU individually and independently. Therefore, an A-MPDU transmission will be considered a complete failure only if all frames in the aggregation are affected by the channel error probability.

	All results are shown with stickiness equal to one (see Section~\ref{errorEffect}). That is, after a colision, CSMA/ECA$_{\text{Hys+FS}}$ nodes will use a random backofff. Nevertheless, in Section~\ref{resultsSchedRest} curves described as \emph{dynStick} temporarily increase the node's stickiness to two after a successful reduction of the schedule (using a random backoff after two consecutive collisions). This increase is done in order to converge faster to collision-free schedules when in operating with channel errors.

@@

	\subsection{Applying Schedule Reset} 
	A set of results under saturated conditions and channel errors applying Schedule Reset (see Section~\ref{schedReset}) are provided. Some of the results are generated with a $\gamma=1$, or \emph{aggressive} Schedule Reset (aggr. in the figures). These settings provide the highest throughput under the tested conditions, and also in the real hardware implementations such as the one shown in Section~\ref{EDCA}.

@@

	This increase in collisions may also cause the dropping of packets due to reaching the maximum retransmission limit. As CSMA/ECA$_{\text{Hys+FS}}$ drops more packets after reaching the retransmission limit (see line~\ref{discard} in Algorithm~\ref{alg:fullECA}), it shows a higher fraction of dropped packets in Figure~\ref{fig:unsatResults}d.

@@

	\begin{figure}[htb!]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated-TON.eps}
		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers are discarded (or blocked) at the entrance of the buffer}
		\label{fig:blocked-packets}
	\end{figure}

@@

@@

	Still assessing Figure~\ref{fig:SchedResetResults}a and focusing on the SR curves, we can see that both aggressive schedule halving with dynamic stickiness (\emph{SR aggr. halv. dynStick}) curves (with and without Fair Share) have higher aggregated throughput than the rest (for a description of aggressive schedule halving and dynamic stickiness, refer to Algorithm~\ref{alg:schedRest} and Section~\ref{aggr}). This Schedule Reset configuration reduces the time between successful transmissions (see Figure~\ref{fig:SchedResetResults}c) and mantains collision-free operation for longer periods of time thanks to the increase in the node's stickiness.

@@

	Figure~\ref{fig:SchedResetResults}b and Figure~\ref{fig:SchedResetResults}d show the aggregated througput and time between successful transmissions, respectively for a non-saturated network with the same $p_e=0.1$. Since under non-saturated conditions CSMA/ECA$_{\text{Hys+FS}}$ nodes reset their schedule every time they empty their MAC queues, Schedule Reset's benefits are not relevant. Nevertheless, a reduction in the time between successful transmissions is observed for Schedule Reset curves in Figure~\ref{fig:SchedResetResults}d.

	%\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-TON.eps}
%		\caption{Average throughput for a saturated network with channel error ($p_e=0.1$). Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}). \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
%		\label{fig:throughput-sat-SR}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/timeBetSxTx-sat-SR-TON.eps}
%		\caption{Time between successful transmissions for a saturated network with channel error ($p_e=0.1$). Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}). \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
%		\label{fig:timeBetSxTx-sat-SR}
%	\end{figure}

\section{Real hardware implementation}\label{EDCA}

@@

As CSMA/ECA$_{\text{Hys}}$ works with the standard 802.11 PHY, we do not need the flexibility offered by complex development kits based on FPGAs like the WARP boards~\cite{amiri2007warp}. Any architecture that allows a customisation of the channel access delay would fit our requirements. For this reason we chose a widely available 802.11b/g WNIC from Broadcom. Like many other cards, its behaviour is ruled by a firmware code that controls the underlying hardware (including the radio, carrier sense and FIFOs) in real time. In particular we used the BRCM4318KBFG PCI card as it is supported by the OpenFWWF~\cite{OpenFWWF} open-source firmware, an alternative to the original code from the manufacturer that has been recently considered as development platform in several research projects~\cite{WMP,gringolitmc14,gringoliccr14,berger14,CF-MAC}.

@@

%\textcolor{black}{\it maybe we could add a figure to explain of the bitmap is filled}
%
%\textcolor{black}{The new section should assume that the implementation has already been done. Further, it should overview:}
%
%\textcolor{black}{\begin{itemize}
%	\item Why do we need it?
%	\item Are there other alternatives?
%	\item How does it do what it does? (I think the explanation of the CF-MAC paper should suffice).
%	\item You can refer to the firmware implementation as CSMA/ECA$_{\text{Hys}}$ with Schedule Halving and dynamic Stickiness. I will define it before this section.
%\end{itemize}}
%
%\textcolor{black}{Obviously this will be a first revision. I leave the rest of the section untouched, we will erase/modify it later. Thanks in advance.}
%
%The implementation of CSMA/ECA$_{\text{Hys+FS}}$ should be carried out at a firmware level due to the tight timing constrains related to the backoff procedure. There are several alternatives to modify the backoff operation in WNICs, either using open firmware like OpenFWWF~\cite{OpenFWWF} or through Field Programmable Gate Arrays (FPGA).
%
%Although the FPGA option would in theory provide the strict timing requirements, the associated costs and difficulty are greater than the OpenFWWF alternative. OpenFWWF provides an open CSMA/CA firmware for a specific model of Broadcom WNICs' chipset, so the resulting firmware can be uploaded and tested in real commercial hardware. 
%
%By making the modifications represented in Algorithm~\ref{alg:CSMA_ECA}, a proof-of-concept prototype of CSMA/ECA was built using comercial hardware~\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test}. The prototype was able to create a collision-free schedule among a reduced number of contenders using big deterministic backoffs (such as $128$ or $256$ slots). Nevertheless, clock drift, imprecise timing~\cite{bianchi2007experimental} and other prototyping non-idialities prevented the construction of a collision-free schedule when using a small deterministic backoff or dynamic rate adaptation algorithms, like Minstrel~\cite{minstrel}.
%
%Although the timing inaccuracy issues were identified and leveraged to prototype a similar protocol in real hardware~\cite{CF-MAC}, further development on the prototype is being carried out in order to incorporate Hysteresis and Fair Share.\textcolor{black}{\section{Transitioning towards CSMA/ECA$_{\text{Hys+FS}}$}\label{ECAtoCA}}
%Given the benefits CSMA/ECA$_{\text{Hys+FS}}$ is able to provide in the tested scenarios, this section aims at summarising the missing features that would allow a transition to CSMA/ECA$_{\text{Hys+FS}}$.

@@

The current PHY/MAC enhancements considered by the HEW Task Group seek to achieve higher throughput and include improved coding schemes; full duplex radios, capable of receiving and transmitting at the same time; OFDMA and Multiple-User Multiple-Input Multiple-Output (MU-MIMO), allowing the transmission of different packets to/from multiple destinations at the same time; as well as dynamic channel bonding techniques~\cite{BorisChannelBonding}. Using CSMA/ECAHys+FS alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decreasing the time spent recovering from collisions.

>>>>

\begin{itemize}
	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
	\item The impact of imperfect clocks and channel errors on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.
	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
	\item Introduce the Schedule Reset mechanism for reducing the schedule length in case of users withdrawing from the contention (non-saturated traffic), or when its length is increased due to channel errors.
	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
	\item First implementation of CSMA/ECA with Hysteresis in real hardware and the experimental results in a crowded WLAN testbed.
\end{itemize}

@@

%
%\input{relatedWork}
%\input{introProtocol}
%\input{simulations}
%\input{results}
%\input{realImplementation-Gringoli}
%\input{transition}

@@

Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like LTE~\cite{canoLTEcoexistence}. Applying the same concept to CSMA networks by modifying DCF's random backoff proceedure provides similar benefits~\cite{HE}. The following are MAC protocols for WLANs, decentralised and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques. A survey of collision-free MAC protocols for WLANs is presented in~\cite{L_MAC}. In this paper we only overview those that are similar to CSMA/ECA.

\subsection{Zero Collision MAC}\label{ZC-MAC}

@@

ZC-MAC is able to outperform CSMA/CA under different scenarios. Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $N>M$. Furthermore, if $M$ is overestimated ($M\gg N$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance. Additionally, ZC-MAC nodes require common knowledge of where the virtual schedule starts/ends. This is not considered in CSMA/CA and constitutes an obstacle towards standardisation.

@@

\begin{figure*}[tb]
\centering
  \includegraphics[width=0.8\linewidth]{figures/basicECA.eps}
  \caption{An example of the temporal evolution of CSMA/ECA in saturation. Slots containing transmissions are around 1000 times larger than empty slots}
  \label{fig:BECA-example}
\end{figure*}

@@

Collision slots being orders of magnitude larger than empty slots degrade the network performance. When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's. Figure~\ref{fig:BECA}, shows the average aggregated throughput of CSMA/ECA, CSMA/CA and collision-free protocolos presented in Section~\ref{relatedWork}, namely ZC-MAC and L-MAC~\cite{L_MAC}.

@@

Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders is greater than $B_{\text{d}}+1$ ($N=8$ in the case of the figure). Beyond this point, the network will have a mixed behaviour relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff. As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's. This effect is also observed in ZC-MAC and L-MAC. It happens becasue the virtual cicle used by these protocolos ($M$ for ZC-MAC and $C$ for L-MAC in Section~\ref{relatedWork}) in lower than the number of contenders, $N$\footnote{L-MAC and ZC-MAC curves in Figure~\ref{fig:BECA} appear to yield the same aggregated throughput. This is because these protocols do not consider an increase in the backoff stage after a failed transmission, augmenting the collision probability beyond CSMA/CA's. An adaptation of the schedule length leverages this issue, which is preseted in~\cite{L_MAC} as \emph{Adaptive} L-MAC and L-ZC.}.

@@

	Channel errors, hidden-nodes, or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations' backoff stages to the maximum value very quickly (contributing to the delay increase). This issue is leveraged with the Schedule Reset mechanim, introduced in Section~\ref{schedReset}.

@@

	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
	Fair Share is an A-MPDU aggregation mechanism~\cite{A-MPDU} that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness. However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads. Furthermore, the level of aggregation provided by Fair Share depends only on the buffer occupancy and the current backoff stage.

@@

	\begin{equation}
			S =
				\begin{cases}\label{eq:bound}
					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{\text{d}}+1 \\
					Ns_{k}(l_{k}), & \text{otherwise}
				\end{cases}
	\end{equation}

@@

	\subsection{Hidden terminals}\label{hidden}
	One key characteristic of IEEE 802.11 devices is that their carrier sense range is at least two times greater than their data range~\cite{tuningCarrierSense}. In this situation, the impact of hidden nodes can be considered to be low. This is because a given transmission could only be interfered by other transmissions from very distant nodes with energy levels not higher than the noise floor, which using modern radios can be easily solved by the capture effect. However, in specific deployments, where obstacles play also an important role on the propagation effects, hidden nodes may appear, causing asynchronous packet collisions~\cite{throughputUnderHiddenTerm}.

	In the case of CSMA/ECA$_{\text{Hys+FS}}$ WLANs, the same negative effects as in IEEE 802.11 WLANs are expected. In addition, collisions with hidden terminals will cause nodes to leave any collision-free schedule, as channel errors do, thus continuously disrupting any attempt of collision-free operation.

@@

	Figure~\ref{fig:scheduleReset1} shows the slots between the transmissions of a contender with $B_{\text{d}}=31$. Starting from the left, the current $B_{\text{d}}=0$ means that the first slot will be filled with the contender's own transmission. Each following slot containing either a transmission or a collision is identified with the number one, while empty slots are marked with a zero. Notice that the highlighted empty slots appear every eight slots, suggesting that a schedule reduction from $B_{\text{d}}=31$ to $B_{\text{d}}^{*}=7$ is possible; where $B_{\text{d}}^{*}$ is the new deterministic backoff assigned by Schedule Reset\footnote{$B_{\text{d}}=2^{0}CW_{\min}/2-1=7$ (see Table~\ref{tab:mac-params}). So the change is made to the backoff stage, $k$. In this case from $2$, to $k=0$. This means that any new schedule must also be a power of two.}. The schedule change is performed after the contender's next successful transmission.

	%, coupled with an increase by one in the node's \emph{stickiness}~\cite{L_MAC2}. 
	%\footnote{Trailing waiting periods, like the SIFS, are composed of empty slots. These allow nodes to decrement their backoff counters even after a collision or successful slots. Without loss of generality, the stations in our examples and simulations decrement their respective backoff counters by one after these kind of slots (see example in Figures~\ref{fig:ECA} and~\ref{fig:ecaQoS}).}
		\begin{figure}[tb]
		\centering
			\includegraphics[width=\linewidth]{figures/tonFigs/scheduleReset.pdf}
			\caption{Example of the Schedule Reset mechanism. The lower row shows values of $t$ for a bitmap $b[t]$ of size $|b[t]| = B_{\text{d}} + 1=32$}
			\label{fig:scheduleReset1}
		\end{figure}

			\begin{algorithm}[tb]
		\If{sxTx == 1}{
			Initialising\;
			\If{B == 0}{
				$b[B_{\text{d}}+1] = \{0\}$\;
				$t = 0$\;
			}
		}
		\If{(sxTx $>$ 0) \&\& (sxTx $\leq~\gamma$)}{
			Filling the bitmap\;
			$b[t] \|\sigma_{i}(t)$\label{bitwiseOR}\;
			$t++$\;
			\If{$t == B_{\text{d}}$}{
				$t=0$\;
			}
		}
		\If{sxTx == $\gamma$}{\label{evaluations}
			Analising\;
			sxTx = $0$\;
			\For{($j=0;~j<k;~j++$)}{\label{sets}
				$y=\lceil 2^{j}CW_{\min}/2\rceil$\;
				isItPossible $= 0$\;
				\For{($m=y;~m<B_{\text{d}}+1;~m+=y$)}{
					isItPossible$~+=b[m]$\;\label{filling}
				}
				\If{isItPossible == $0$}{\label{testFilling}
					change = $1$\;
					newStage = $m$\;
					{\bf break};
				}
			}
		}
		\If{change == $1$}{
			Making the change\;
			$k$ = newStage\;
			change = $0$\;\label{endAssignment}
		}
		\vspace{0.2cm}
		\caption{Schedule Reset Mechanism for CSMA/ECA$_{\text{Hys+FS}}$. Every consecutive successful transmission increases the variable SxTx by one, while a collision resets it to zero. The algorithm is called a \emph{reset} when all smaller deterministic backoff are tested. On the other hand, when $j$ in line~\ref{sets} is initialised to $j=k-1$, it is called a schedule \emph{halving}}\label{alg:schedRest}
	\end{algorithm}

@@

	\subsection{Channel errors}\label{channelErrorsDef}
	Channel errors are modeled by assigning a probability of a packet being corrupted by the channel, $p_e>0$. That is, in a single packet transmission there is probability $p_e$ that the transmission will not be acknowledged. If the transmission is an A-MPDU (like in the case of CSMA/ECA$_{\text{Hys+FS}}$), $p_e$ will affect each MPDU individually and independently. Therefore, an A-MPDU transmission will be considered a complete failure only if all frames in the aggregation are affected by the channel error probability.

	All results are shown with stickiness equal to one (see Section~\ref{errorEffect}). That is, after a colision, CSMA/ECA$_{\text{Hys+FS}}$ nodes will use a random backofff. Nevertheless, in Section~\ref{resultsSchedRest} curves described as \emph{dynStick} temporarily increase the node's stickiness to two after a successful reduction of the schedule (using a random backoff after two consecutive collisions). This increase is done in order to converge faster to collision-free schedules when operating with channel errors.

@@

	\subsection{Applying Schedule Reset} 
	A set of results under saturated conditions and channel errors applying Schedule Reset (see Section~\ref{schedReset}) are provided. Some of the results are generated with a $\gamma=1$, or \emph{aggressive} Schedule Reset (indicated as \emph{aggr.} in the figures). These settings provide the highest throughput under the tested conditions, and also in the real hardware implementations such as the one shown in Section~\ref{EDCA}.

@@

	This increase in collisions may also cause the dropping of packets due to reaching the maximum retransmission limit. As CSMA/ECA$_{\text{Hys+FS}}$ drops more packets after reaching such limit (see line~\ref{discard} in Algorithm~\ref{alg:fullECA}), it shows a higher fraction of dropped packets in Figure~\ref{fig:unsatResults}d.

@@

@@

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated-TON.eps}
		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers are discarded (or blocked) at the entrance of the buffer}
		\label{fig:blocked-packets}
	\end{figure}

@@

	Still assessing Figure~\ref{fig:SchedResetResults}a and focusing on the SR curves, we can see that both aggressive schedule halving with dynamic stickiness (\emph{SR aggr. halv. dynStick}) curves (with and without Fair Share) have higher aggregated throughput than the rest (for a description of aggressive schedule halving and dynamic stickiness, refer to Algorithm~\ref{alg:schedRest} and Section~\ref{aggr}). This Schedule Reset configuration reduces the time between successful transmissions (see Figure~\ref{fig:SchedResetResults}c) and maintains collision-free operation for longer periods of time thanks to the increase in the node's stickiness.

@@

	Figure~\ref{fig:SchedResetResults}b and Figure~\ref{fig:SchedResetResults}d show the aggregated throughput and time between successful transmissions, respectively for a non-saturated network with the same $p_e=0.1$. Since under non-saturated conditions CSMA/ECA$_{\text{Hys+FS}}$ nodes reset their schedule every time they empty their MAC queues, Schedule Reset's benefits are not relevant. Nevertheless, a reduction in the time between successful transmissions is observed for Schedule Reset curves in Figure~\ref{fig:SchedResetResults}d.

	%\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-TON.eps}
%		\caption{Average throughput for a saturated network with channel error ($p_e=0.1$). Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}). \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
%		\label{fig:throughput-sat-SR}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/timeBetSxTx-sat-SR-TON.eps}
%		\caption{Time between successful transmissions for a saturated network with channel error ($p_e=0.1$). Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}). \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
%		\label{fig:timeBetSxTx-sat-SR}
%	\end{figure}
\section{Real hardware implementation}\label{EDCA}

@@

As CSMA/ECA$_{\text{Hys}}$ works with the standard 802.11 PHY, we do not need the flexibility offered by complex development kits based on FPGAs like the WARP boards~\cite{amiri2007warp}. Any architecture that allows a customisation of the channel access delay{black} would fit our requirements. For this reason we chose a widely available 802.11b/g WNIC from Broadcom. Like many other cards, its behaviour is ruled by a firmware code that controls the underlying hardware (including the radio, carrier sense and FIFOs) in real time. In particular we used the BRCM4318KBFG PCI card as it is supported by the OpenFWWF~\cite{OpenFWWF} open-source firmware, an alternative to the original code from the manufacturer that has been recently considered as development platform in several research projects~\cite{WMP,gringolitmc14,gringoliccr14,berger14,CF-MAC}.

@@

%\textcolor{black}{\it maybe we could add a figure to explain of the bitmap is filled}
%
%\textcolor{black}{The new section should assume that the implementation has already been done. Further, it should overview:}
%
%\textcolor{black}{\begin{itemize}
%	\item Why do we need it?
%	\item Are there other alternatives?
%	\item How does it do what it does? (I think the explanation of the CF-MAC paper should suffice).
%	\item You can refer to the firmware implementation as CSMA/ECA$_{\text{Hys}}$ with Schedule Halving and dynamic Stickiness. I will define it before this section.
%\end{itemize}}
%
%\textcolor{black}{Obviously this will be a first revision. I leave the rest of the section untouched, we will erase/modify it later. Thanks in advance.}
%
%The implementation of CSMA/ECA$_{\text{Hys+FS}}$ should be carried out at a firmware level due to the tight timing constrains related to the backoff procedure. There are several alternatives to modify the backoff operation in WNICs, either using open firmware like OpenFWWF~\cite{OpenFWWF} or through Field Programmable Gate Arrays (FPGA).
%
%Although the FPGA option would in theory provide the strict timing requirements, the associated costs and difficulty are greater than the OpenFWWF alternative. OpenFWWF provides an open CSMA/CA firmware for a specific model of Broadcom WNICs' chipset, so the resulting firmware can be uploaded and tested in real commercial hardware. 
%
%By making the modifications represented in Algorithm~\ref{alg:CSMA_ECA}, a proof-of-concept prototype of CSMA/ECA was built using comercial hardware~\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test}. The prototype was able to create a collision-free schedule among a reduced number of contenders using big deterministic backoffs (such as $128$ or $256$ slots). Nevertheless, clock drift, imprecise timing~\cite{bianchi2007experimental} and other prototyping non-idialities prevented the construction of a collision-free schedule when using a small deterministic backoff or dynamic rate adaptation algorithms, like Minstrel~\cite{minstrel}.
%
%Although the timing inaccuracy issues were identified and leveraged to prototype a similar protocol in real hardware~\cite{CF-MAC}, further development on the prototype is being carried out in order to incorporate Hysteresis and Fair Share.

@@

%Given the benefits CSMA/ECA$_{\text{Hys+FS}}$ is able to provide in the tested scenarios, this section aims at summarising the missing features that would allow a transition to CSMA/ECA$_{\text{Hys+FS}}$.

The current PHY/MAC enhancements considered by the HEW Task Group seek to achieve higher throughput and include improved coding schemes; full duplex radios, capable of receiving and transmitting at the same time; OFDMA and Multiple-User Multiple-Input Multiple-Output (MU-MIMO), allowing the transmission of different packets to/from multiple destinations at the same time; as well as dynamic channel bonding techniques~\cite{BorisChannelBonding}. Using CSMA/ECA$_{\text{Hys+FS}}$ alongside these features would provide enhanced performance by constructing collision-free schedules, thus substantially decre{black}asing the time spent recovering from collisions.

### 2015-11-09 relatedWork.tex
Finished

\textcolor{red}{Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like LTE~\cite{canoLTEcoexistence}. Applying the same concept to CSMA networks by modifying DCF's random backoff proceedure provides similar benefits~\cite{HE}. The following are MAC protocols for WLANs, decentralised and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques. A survey of collisio-free MAC protocols for WLANs is presented in~\cite{L_MAC}. In this paper we only overview those that are similar to CSMA/ECA.}

\subsection{Zero Collision MAC}

>>>>

\textcolor{black}{Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like LTE~\cite{canoLTEcoexistence}. Applying the same concept to CSMA networks by modifying DCF's random backoff proceedure provides similar benefits~\cite{HE}. The following are MAC protocols for WLANs, decentralised and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques. A survey of collision-free MAC protocols for WLANs is presented in~\cite{L_MAC}. In this paper we only overview those that are similar to CSMA/ECA.}

\subsection{Zero Collision MAC}\label{ZC-MAC}
